---
title: "Complete Guide to ggplot2 Geoms"
author: "Gabby Novak"
date: '2019-08-14'
draft: true
tags: ["ggplot2"]
categories: ["Graphics"]
---

```{r setup, include=F}
knitr::opts_chunk$set(collapse = TRUE, 
                      fig.height=5, 
                      fig.width=5,
                      message=F,
                      warning=F,
                      error=F,
                      echo=F,
                      fig.align="center")
library(gridExtra)
library(ggplot2)
library(knitr)
library(kableExtra)
library(magrittr)
set.seed(08142019)
data<-data.frame(x=sample(1:200,100),
                 y=sample(1:200,100))


```

ggplot graphics are built by what are called geoms. These basically determine that type of graph that will be produced. Several geoms can be layered into a single graphic, each using different data if necessary. This will be a basic run down of the availible geoms and their common uses.

# Reference Lines

```{r, fig.width=8,fig.height=2.5}
grid.arrange(
  ggplot(data,aes(x,y))+geom_blank()+geom_abline(intercept=0,slope=1)+labs(title="geom_abline()"),
  ggplot(data,aes(x,y))+geom_blank()+geom_hline(yintercept=100)+labs(title="geom_hline()"),
  ggplot(data,aes(x,y))+geom_blank()+geom_vline(xintercept=100)+labs(title="geom_vline()"),
  ncol=3)

data.frame(Geom=c("geom_abline()","geom_hline()","geom_vline()"),
           Data=c("slope; intercept","yintercept","xintercept"),
           Purpose=c("Diagonal reference line","Horizontal reference line","Vertical reference line"))%>%
  kable(bootstrap=T,align="c")%>%
  kable_styling(bootstrap_options = c("striped"), full_width = T,position="center")
```

# Bar Charts

Coming soon

# Heatmaps

Coming soon

# Blank plot

Coming soon

# Box plot

Coming soon

# 2d contours of a 3d surface

Coming soon

# Count overlapping points

Coming soon

# Smoothed Density Estimates

Coming soon

# Contours of a 2d density estimate

Coming soon

# Dot plot

Coming soon

# Horizontal error bars

Coming soon

# Histograms

Coming soon

# Jittered points

Coming soon

# Vertical intervals

Coming soon

# Polygons from a reference map

Coming soon

# Connecting points

Coming soon

# Points

Coming soon

# Ploygons

Coming soon

# Quantile-quantile plot

Coming soon

# Quantile regression

Coming soon

# Ribbons

Coming soon

# Rug plots

Coming soon

# Line segments

Coming soon

# Smoothed conditional means

Coming soon

# Text

Coming soon

# Rectangles

Coming soon

# Violin plots

Coming soon

# sf objects

Coming soon


**[Full ggplot2 documentation](https://ggplot2.tidyverse.org/reference/)**