---
title: "Subsetting"
author: "Gabby Novak"
date: '2019-08-09'
subtitle: "The Practical Bits"
tags: ["Base R"]
categories: ["R for Rookies"]
---

```{r setup, include=F}
knitr::opts_chunk$set(collapse = TRUE, 
                      fig.height=5, 
                      fig.width=5,
                      message=F,
                      warning=F,
                      error=F,
                      fig.align="center")
# dir<-"C:/Users/gabri/Documents/R/R4Humans-AuxFiles/MappingWithggplot2"
```

# Subsetting in 1 Dimention

Subsetting is simply the act of taking an object and extracting elements from it in some controlled manner. If the subset is assigned to a variable then it can be used just as the original object was. Here we will discuss subsetting in 1 and 2 dimentions (vectors and data frames). Although higher dimension objects do exist in R, the methods for 2 dimentional objects are easily generalizable.

```
General form for subsetting in 1 dimention:
<object>[<SubsettingCondition>]
```

```{r}
# The vector we will be subsetting
x <- c(1,1,2,3,5,8,13)

# Keep in mind that vectors cannot mix data types
# The code below creates a character vector; the numbers are coerced into characters
# x <- c(1,2,3,"a","b","c")
```

## Subsetting by Indexing

R, like other programming languages, assigns indexes to objects containing multiple elements. An index is simply a numerical reference, based on the element's position in the order of the object, beginning at 1 for the first element and continuing sequentially until the last element. Subsetting by referencing these positions is perhaps the most straight forward method.

```{r}
# Subsetting the first element from x
x[1]

# Subsetting the second element from x
x[2]

# Subsetting the 7th element from x
x[7]

# Attempting to subset outside the range of the object (there is no 8th element in x) returns NA
x[8]

# Subsetting elements 3 through 6 from x
x[3:6]
```

